Global Life Expectancy: Unraveling Health and Economic Determinants (2015-2020)

### 2024-12-01 paper/paper.qmd
initial

# load data
lelecleaned_data<- read_parquet("data/02-analysis_data/lelecleaned.parquet")
lea60 <- read_parquet("data/02-analysis_data/lea60.parquet")
leab <- read_parquet("data/02-analysis_data/leab.parquet")
male_table <- read_parquet("data/02-analysis_data/male_table.parquet")
female_table <- read_parquet("data/02-analysis_data/female_table.parquet")
```

@@

The primary goal of this paper is to determine which factors play a statistically significant role in driving lower life expectancy values and to offer actionable insights based on country status. Using Multiple Linear Regression model, this study focuses on understanding the relationship between country status(developed or developing), personal average expenditure on health based on GDP, percentage of government expenditure on health, income composition of resources and year of schooling in predicting life expectancy across different countries. By identifying and analyzing these predictors, the study aims to highlight actionable aspects for policymakers to target in their efforts to improve population longevity effectively.

Related Research has shown that higher healthcare expenditure of government positively impact life expectancy [@valueofvaccination]. Additionally, addressing social determinants such as education and income inequality has been identified as a vital pathway to improving longevity [@healthequity]. Studies also highlight the role of global collaboration in tackling health crises, which can disproportionately affect lower-income nations, mostly developing countries under certain continents, further influencing life expectancy disparities [@World_Health_Organization].

Findings from this paper reveal that countries with greater healthcare spending tend to achieve higher life expectancy, while factors like income inequality and absence of education emerge as significant obstacles. The analysis underscores the importance of targeted interventions in key areas such as healthcare accessibility and economic equity to address disparities in life expectancy. This study further contributes to a deeper understanding of how predictive modeling under different country status can inform public health strategies and policy-making on a global scale.

>>>>

# load data
lelecleaned_data <- read_parquet("data/02-analysis_data/lelecleaned.parquet")
lea60 <- read_parquet("data/02-analysis_data/lea60.parquet")
leab <- read_parquet("data/02-analysis_data/leab.parquet")
male_table <- read_parquet("data/02-analysis_data/male_table.parquet")
female_table <- read_parquet("data/02-analysis_data/female_table.parquet")
```

@@

The primary goal of this paper is to determine which factors play a statistically significant role in driving lower life expectancy values and to offer actionable insights based on different income group. Using Multiple Linear Regression model and Bayesian model, this study focuses on understanding the relationship between Gender, Region and different income group in predicting life expectancy across different countries. By identifying and analyzing these predictors, the study aims to highlight actionable aspects for policymakers to target in their efforts to improve population longevity effectively.

Related Research has shown that there exist difference between men and women related with biological, behavioral, and socioeconomic factors, highlighting that gender-specific health behaviors and societal roles influence longevity [@GenderandLifeExpectancy]. Higher-income individuals tend to live longer due to better access to healthcare and healthier lifestyles, with notable regional disparities even within similar income groups [@IncomeGroupsandLongevity]. Additionally, regional socioeconomic differences on health outcomes, especially life expectancy. It emphasizes that addressing regional inequities in wealth and resources is essential for improving population health globally. [@RegionalVariations].

Findings from this paper reveal that countries of higher income group tend to achieve higher life expectancy, while factors like lower income and poverty region emerge as significant obstacles. The analysis underscores the importance of targeted interventions in key areas such as healthcare accessibility and economic equity to address disparities in life expectancy. This study further contributes to a deeper understanding of how predictive modeling under different country status can inform public health strategies and policy-making on a global scale.
